%!TEX root = ../thesis.tex
%*******************************************************************************
%****************************** Third Chapter **********************************
%*******************************************************************************
\chapter{Protocols and analysis of single-cell DNA methylation data} \label{sec:pro}

% **************************** Define Graphics Path **************************
\ifpdf
    \graphicspath{{Chapter3/Figs/Raster/}{Chapter3/Figs/PDF/}{Chapter3/Figs/}}
\else
    \graphicspath{{Chapter3/Figs/Vector/}{Chapter3/Figs/}}
\fi

Our understanding of DNA methylation has been revolutionized by the development of BS-seq, which offers single-cytosine resolution and absolute quantification of 5mC genome-wide. Recent advances have demonstrated the power of single-cell sequencing to deconvolve mixed cell populations~\citep{jain_supervised_2007,deng_single-cell_2014,macaulay_single_2014}. Incorporating epigenetic information into this single-cell arsenal will provide insights into epigenetic heterogeneity and broaden our understanding of gene regulation.

In the first section of this chapter, we will describe the scBS-seq protocol for genome-wide profiling of DNA methylation in single cells, a statistical method for quantifying methylation variability between cells, and applications to mouse ESCs. The presented work is based on \citet{smallwood_single-cell_2014}, which was joint work of Sebastien Smallwood, Heather Lee, Christof Angermueller, Felix Krueger, Heba Saadeh, Julian Peat, Simon Andrews, Oliver Stegle, and Wolf Reik.

\begin{center}
\begin{minipage}{.9\linewidth}
\underline{Individual contributions}: Sebastien Smallwood and Heather Lee designed the study, prepared scBS-seq libraries, analysed data and wrote the manuscript. Felix Krueger, Heba Saadeh, and Sebastien Smallwood performed sequence mapping and analysed data. Julian Peat contributed to technical developments. Christof Angermueller and Oliver Stegle analysed the data.
\end{minipage}
\end{center}

In the second section, we will describe the scM\&T-seq protocol for parallel profiling of DNA methylation and gene expression in single cells, methods for quantifying associations between DNA methylation and gene expression, and applications to mouse ESCs. The presented work is based on \citet{angermueller_parallel_2016}, which was joint work of Christof Angermueller, Stephen Clark, Heather Lee, Iain Macaulay, Mabel Teng, Tim Xiaoming Hu, Felix Krueger, Sebastien Smallwood, Chris Ponting, Thierry Voet, Gavin Kelsey, Oliver Stegle, and Wolf Reik.

\begin{center}
\begin{minipage}{.9\linewidth}
\underline{Individual contributions}: Christof Angermueller performed all statistical analyses of the data. Heather Lee, Iain Macaulay, Stephen Clark, and Sebastien Smallwood developed the protocol and performed experiments. Heather Lee, Iain Macaulay, Christof Angermueller, Stephen Clark, Oliver Stegle, Wolf Reik, and Chris Ponting interpreted the results. Mabel Teng contributed to method development. Tim Xiaoming Hu processed RNA-seq data. Felix Krueger processed BS-seq data. Wolf Reik, Gavin Kelsey, Iain Macaula, and Thierry Voet contributed protocols and reagents. Heather Lee, Iain Macaulay, Wolf Reik, and Thierry Voet conceived the project.
\end{minipage}
\end{center}

\input{./Chapter3/bs/main.tex}
\input{./Chapter3/mt/main.tex}


\section{Discussion}

We have first presented scBS-seq for profiling DNA methylation in single cells and a statistical model for quantifying methylation heterogeneity at a genome-wide scale. We demonstrated that our approach detects rare cell types and identifies genomic regions with high methylation heterogeneity that are functionally important during cell differentiation.

We further extended our protocol for parallel profiling of the methylome and transcriptome from the same single cell, thereby allowing to probe regulatory relationships between DNA methylation and expression. We both confirmed previously known negative associations between non-CGI promoter methylation and transcription in single cells, and identified novel positive and negative associations at distal regulatory regions. We found the expression levels of many pluripotency factors, e.g. Esrrb, to be negatively associated with DNA methylation. Furthermore, we found evidence that the strength of associations between methylome and transcriptome can vary from cell to cell.

We used principal component analysis to investigate the factors of variations of DNA methylation and gene expression. To better understand the relation between these factors and to distinguish between factors that drive variations either in DNA methylation or gene expression, future studies may consider advanced statistical methods, such as canonical correlation analysis~\citep{hardoon_canonical_2004,everitt_canonical_2005}, or collective matrix factorization~\citep{singh_relational_2008,bouchard_convex_2013,cao_multi-view_2015}.

In our study, we quantified local associations between genes and genomic regions that were separated by at most 10~kbp. An important area of future research is to also investigate trans-associations beyond 10~kbp. Appealing for this purpose are chromatin contact information from Hi-C experiments~\citep{de_wit_decade_2012}, which can help to identify distal genomic regions that are spatially close to a certain gene.

We estimated mean methylation rates in genomic contexts by averaging and used weighted Pearson correlation to account for incomplete CpG coverage. More sophisticated methods for estimating the mean methylation rate in genomic contexts from low-coverage data may reveal stronger associations between DNA methylation and gene expression. These include methods for imputing CpG methylation in single cells, which will be discussed in the following chapter.
